### 2023-06-06 proofs/tex/alg-pos-subtyping-lemmas.tex
Restructure

\begin{lemma}[Soundness of Positive Subtyping] \label{lemma:pos-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
    implies $[[us : Θ|uv uP]]$ and 
    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
\end{lemma}
\begin{proof} 
    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
    \begin{caseof}
    \item \ruleref{\ottdruleAPUVarLabel} then
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.

@@

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
        implies $[[us : Θ|uv uN]]$ and 
        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uN ≤ iM ]]$
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$.
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item \ruleref{\ottdruleAArrowLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape
        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
        On the next step, the the algorithm makes two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
        By the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}) and induction hypothesis, respectively 
        \begin{enumerate}
            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
        \end{enumerate}

        Then the algorithm merges two unification solutions $[[us1]]$ and $[[us2]]$.
        By \cref{lemma:merge-soundness}, since $[[uv uP ∪ uv uN' = uv (uP → uN')]]$, 
        we have $[[us1 & us2 : Θ | uv (uP → uN')]]$, and also
        $[[Θ ⊢ us1 & us2 ⇒ us1]]$ and $[[Θ ⊢ us1 & us2 ⇒ us2]]$.
        By the transitivity of solution weakening (\cref{lemma:weakening-transitivity}),
         $[[Θ ⊢ us' ⇒ us1 & us2]]$ implies $[[Θ ⊢ us' ⇒ us1]]$ and $[[Θ ⊢ us' ⇒ us2]]$.

% \[\begin{tikzcd}[arrows=Rightarrow]
% 	& {[[us']]} \\
% 	{[[us1]]} & {[[us1 & us2]]} & {[[us2]]}
% 	\arrow[from=2-2, to=2-3]
% 	\arrow[from=2-2, to=2-1]
% 	\arrow[from=1-2, to=2-2]
% \end{tikzcd}\]

        The application of the induction hypothesis gives us 
        $[[Γ ⊢ [us']uP ≥ iQ ]]$ and $[[ Γ ⊢ [us']uN' ≤ iM' ]]$.
        Finally, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [us'](uP → uN') ≤ iQ → iM']]$.

        \item \ruleref{\ottdruleANVarLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
        This case is symmetric to \cref{case:pos-subt-soundness:var} of \cref{lemma:pos-subt-soundness}.

        \item \ruleref{\ottdruleAShiftULabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
        This case is symmetric to \cref{case:pos-subt-soundness:shift} of \cref{lemma:pos-subt-soundness}.

        \item \ruleref{\ottdruleAForallLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
        This case is symmetric to \cref{case:pos-subt-soundness:exists} of \cref{lemma:pos-subt-soundness}.

\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
\end{theorem}
\begin{proof}
\end{proof}

@@

\begin{lemma}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
    $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
    and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
    and $[[Θ ⊢ us ⇒ us']]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [us]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneArrowLabel}, 
        i.e. $[[ [us]uN ]] = [[ [us]uP → [us]uN' ]]$ and $[[iM]] = [[iQ → iM']]$, 
        and by inversion, $[[ Γ ⊢ [us]uP ≥ iQ ]]$ and $[[ Γ ⊢ [us]uN' ≤ iM' ]]$.

        Let us consider restrictions of $[[us]]$ to 
        the set of unification variables in $[[uP]]$ and $[[uN']]$:
        $[[us | uv uP : Θ | uv uP]]$ and $[[us | uv uN' : Θ | uv uN']]$.
        Notice that 
        $[[ [us]uP = [us | uv uP]uP ]]$ and 
        $[[ [us]uN' = [us | uv uN']uN' ]]$.

       Let us apply the induction hypothesis to
       $[[ Γ ⊢ [us | uv uP]uP ≥ iQ ]]$ and
       $[[ Γ ⊢ [us | uv uN']uN' ≤ iM' ]]$ (notice that
       since $[[uv uN' ⊆ uv uN]]$, there are no negative unification variables in $[[uN']]$)
       to obtain $[[us'1]]$ and $[[us'2]]$ such that
       \begin{enumerate}
        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'1]]$ and $[[Θ ⊢ us | uv uP ⇒ us'1]]$
        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'2]]$ and $[[Θ ⊢ us | uv uN' ⇒ us'2]]$ 
       \end{enumerate}

       This way, the algorithm applies \ruleref{\ottdruleAArrowLabel}
       and terminates returning $[[us'1 & us'2]]$ as the result.

       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
       Since $[[us | uv uP ⊆ us]]$, 
       by \cref{lemma:weakening-monotonicity}, 
       $[[Θ ⊢ us ⇒ us | uv uP]]$,
       and then by transitivity (\cref{lemma:weakening-transitivity}), 
       $[[Θ ⊢ us ⇒ us'1]]$.
       Analogously, $[[Θ ⊢ us ⇒ us'2]]$.
       Then since by \cref{lemma:merge-completeness}, 
       $[[us'1 & us'2]]$ is the weakest restriction  
       implying both $[[us'1]]$ and $[[us'2]]$,
       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.

       \item \label{case:subt-complete-forall}
       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel}.
       Since $[[uN]]$ does not contain negative unification variables,
       $[[uN]]$ must be of the form $[[∀pas.uN']]$,
       such that $[[ [us]uN = ∀pas.[us]uN' ]]$ and $[[ [us]uN']] \neq [[∀]]\dots$
       (assuming $[[pas]]$ does not intersect with the range of $[[us]]$).
       Also, $[[iM]] = [[∀pbs.iM']]$ and either $[[pas]]$ or $[[pbs]]$ is non-empty.

       The rest of the proof is symmetric to \cref{case:pos-subt-complete-exists} of
       \cref{lemma:pos-subt-completeness}.
       To apply the induction hypothesis, we need to show additionally that
       there are no negative unification variables in $[[uN0]] = [[ [â⁺*/pas]uN' ]]$.
       This is because $[[ uv uN0 ⊆ uv uN ∪ {â⁺*} ]]$, and $[[uN]]$ is free of negative
       unification variables by assumption.

       \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}.\\
       Then $[[iN]] = [[ [us]uN ]] = [[ α⁻ ]] = [[iM]]$. 
       Here the first equality holds because $[[uN]]$ is not a unification variable:
       by assumption, $[[uN]]$ is free of negative unification variables.
       The second and the third equations hold because \ruleref{\ottdruleDOneNVarLabel}
       was applied.

       The rest of the proof is symmetric to \cref{case:pos-subt-complete-pvar} of
       \cref{lemma:pos-subt-completeness}.

>>>>

\begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
    implies $[[us : Θ|uv uP]]$ and 
    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
\end{lemma}
\begin{proof} 
    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
    \begin{caseof}
    \item \ruleref{\ottdruleAPUVarLabel} then
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.

@@

@@
